Speeding up decimal multiplication

### 2020-11-22 paper/paper.tex
Update

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathrsfs}
\usepackage{cite}
\usepackage{url}
\usepackage[ruled]{algorithm2e}

@@

\begin{abstract}
    Decimal multiplication is the task of multiplying two numbers in base $10^N.$
    Specifically, we focus on the number-theoretic transform (NTT) family of algorithms.
    Using only portable techniques, we achieve a 3x---5x speedup over the \textbf{mpdecimal} library.
    In this paper we describe our implementation and discuss further possible optimizations.
    Our contributions are the following:
    (1) the ``linearity trick'' for using Montgomery reduction with NTT without separate steps of conversion into/out of Montgomery form;
    (2) a simple cache-efficient algorithm for in-place $2n \times n$ or $n \times 2n$ matrix transposition, the need for which arises in the ``six-step algorithm'' variation of the matrix Fourier algorithm, and which does not seem to be widely known.
    Notable findings include:
    (1) use of two prime moduli instead of three makes sense even considering the worst case of increasing the size of the input, and makes for simpler answer recovery;
    (2) the ``six-step algorithm'', supposedly fast for large transforms due to cache-friendliness, is of little use on modern systems for transforms of sizes up to $\approx 2^{27}.$
\end{abstract}

@@

In order to get a provably correct answer, you have to put less bits of information into the
floating-point values than otherwise possible~\cite{gmp_paper, mattcomp_precision}.
Because of that, in order to achieve performance
parity with multiplication algorithms that abstain from use of floating point, platform-specific
SIMD extensions are often used~\cite{gmp_paper}.
Also, in order to get a provably correct result, you need to know the precision of your
floating-point types, as well as possible quirks of their implementation --- i.e., depend on the
platform.

@@

Remember also that the ``high compositeness'' of $N,$ which must divide $\gcd(p_1 - 1, \ldots, p_n - 1),$ is required
for the Cooley-Tukey algorithm to work on a length-$N$ transform. What a coincidence!

@@

This means that, for the purpose of computing the cyclic convolution, we can use DIF for the direct
transform and DIT for the inverse transform, and simultaneously omit the final permutation step in
DIF and the initial permutation step in DIT.
But note this does not mean just one of them can be omitted but not the other!

@@

The latter approach is preferable for us: the Cooley-Tukey IDFT performs multiplication by
scalar $N^{-1}$ as a separate final stage; we can ``mix in'' the factor of $R$ into this stage
essentially for free!

@@

We now describe the four-step and six-step algorithms for DFT/IDFT, which are matrix Fourier
algorithms, meaning that they interpret sequences of length $N = N_1 N_2$ as $N_1 \times N_2$
matrices.

@@

DFT of length $N = RC$ can be computed as follows~\cite{mattcomp_mfa}:
\begin{enumerate}
    \item $[R \times C]$, perform (length $R$) DFT on each column with $\xi = (\xi_N)^C.$
    \item $[R \times C]$, multiply each element indexed $\langle i,j \rangle$ by $(\xi_N)^{ij}.$
    \item $[R \times C]$, perform (length $C$) DFT on each row with $\xi = (\xi_N)^R.$
    \item $[R \times C]$, transpose the matrix.
\end{enumerate}

@@

DFT of length $N = RC$ can be computed as follows~\cite{mattcomp_mfa}:
\begin{enumerate}
    \item $[R \times C],$ transpose the matrix.
    \item $[C \times R],$ perform (length $R$) DFT on each row with $\xi = (\xi_N)^C.$
    \item $[C \times R],$ transpose the matrix.
    \item $[R \times C],$ multiply each element indexed $\langle i,j \rangle$ by $(\xi_N)^{ij}.$
    \item $[R \times C],$ perform (length $C$) DFT on each row with $\xi = (\xi_N)^R.$
    \item $[R \times C],$ transpose the matrix.
\end{enumerate}

@@

@@

It is possible the built-in two-word types: \verb~uint64_t~ on 32-bit systems and
\verb~unsigned __int128~ on 64-bit systems. But the latest versions of both Clang and GNU GCC do no
optimize the division of two-word types by a constant, as they do for one-word types. Thus,
this built-in division is slow.

@@

Note that we can pre-calculate the value of $r.$ We can also pre-calculate
the Montgomery representation (see section~\ref{section:montgomery_intro})
of $r$ for the second modulo,
$$
\widetilde{r} = (R \cdot r) \mymod p_2.
$$
Then, the value of $( (r \cdot \psi) \mymod p_2 )$ can be calculated as
$$\varphi = \redc(\widetilde{r}, \psi),$$
where the $\redc$ is done for the modulo $p_2.$

@@

Remember the procedure $\mathrm{MontgomeryReduce}(a)$ that we defined in
section~\ref{section:montgomery_intro}.

\subsection{Cache aliasing problems}

>>>>

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathrsfs}
\usepackage{cite}
\usepackage{url}
\usepackage[ruled]{algorithm2e}
\usepackage{graphicx,xcolor}
\usepackage{gnuplottex}
\usepackage{float}

@@

\begin{abstract}
    Decimal multiplication is the task of multiplying two numbers in base $10^N.$
    Specifically, we focus on the number-theoretic transform (NTT) family of algorithms.
    Using only portable techniques, we achieve a 3x---5x speedup over the \textbf{mpdecimal}
    library.
    In this paper we describe our implementation and discuss further possible optimizations.
    We also present a simple cache-efficient algorithm for in-place $2n \times n$ or
    $n \times 2n$ matrix transposition, the need for which arises in the ``six-step algorithm''
    variation of the matrix Fourier algorithm, and which does not seem to be widely known.
    Another finding is that use of two prime moduli instead of three makes sense even considering
    the worst case of increasing the size of the input, and makes for simpler answer recovery.
\end{abstract}

@@

In order to get a provably correct answer, you have to put less bits of information into the
floating-point values than otherwise possible
(tables of maximum number of decimal/binary digits are given in
\cite[p. 560--561]{mattcomp} and \cite{gmp_paper};
for explicit formulas, see \cite{precision_paper}).
Because of that, in order to achieve performance
parity with multiplication algorithms that abstain from use of floating point, platform-specific
SIMD extensions are often used~\cite{gmp_paper}.
Also, in order to get a provably correct result, you need to know the precision of your
floating-point types, as well as possible quirks of their implementation --- i.e., depend on the
platform.

@@

Remember also that the ``high compositeness'' of $N,$ which must divide $\gcd(p_1 - 1, \ldots, p_n - 1),$ is required
for the Cooley-Tukey algorithm to work on a length-$N$ transform. Note the coincidence!

@@

This means that, for the purpose of computing the cyclic convolution, we can use DIF for the direct
transform and DIT for the inverse transform, and simultaneously omit the final permutation step in
DIF and the initial permutation step in DIT.
Note this does not mean just one of them can be omitted but not the other.

@@

The latter approach is preferable for us: the Cooley-Tukey IDFT performs multiplication by
scalar $N^{-1}$ as a separate final stage; we can ``mix in'' the factor of $R$ into this stage
essentially for free.

@@

We now describe the four-step and six-step algorithms for DFT/IDFT, which are matrix Fourier
algorithms, meaning that they interpret sequences of length $N = N_1 N_2$ as $N_1 \times N_2$
matrices~\cite[p. 438--439]{mattcomp}.

@@

DFT of length $N = RC$ can be computed as follows:
\begin{enumerate}
    \item $[R \times C]$, perform (length $R$) DFT on each column with $\xi = (\xi_N)^C.$
    \item $[R \times C]$, multiply each element indexed $\langle i,j \rangle$ by $(\xi_N)^{ij}.$
    \item $[R \times C]$, perform (length $C$) DFT on each row with $\xi = (\xi_N)^R.$
    \item $[R \times C]$, transpose the matrix.
\end{enumerate}

@@

DFT of length $N = RC$ can be computed as follows:
\begin{enumerate}
    \item $[R \times C],$ transpose the matrix.
    \item $[C \times R],$ perform (length $R$) DFT on each row with $\xi = (\xi_N)^C.$
    \item $[C \times R],$ transpose the matrix.
    \item $[R \times C],$ multiply each element indexed $\langle i,j \rangle$ by $(\xi_N)^{ij}.$
    \item $[R \times C],$ perform (length $C$) DFT on each row with $\xi = (\xi_N)^R.$
    \item $[R \times C],$ transpose the matrix.
\end{enumerate}

@@

\subsection{Algorithm for bit-reversal permutation}

See~\cite{bit_rev_perm} for overview of algorithms for bit-reversal permutation.
We use the ``XOR'' approach presented therein, which is simple and has competitive performance
for small-to-medium sequence sizes.

@@

It is possible to use the built-in two-word types, \verb~uint64_t~ on 32-bit systems and
\verb~unsigned __int128~ on 64-bit systems, for this division;
but neither the latest Clang nor GNU GCC do optimize the division of two-word types by a
constant, as they both do for one-word types. Thus, this built-in division is slow.

@@

Note that we can pre-calculate the value of $r.$ We can also pre-calculate
the Montgomery representation (see section~\ref{section:montgomery_intro})
of $r$ for the second modulo,
$$
\widetilde{r} = (R \cdot r) \mymod p_2,
$$
where $R = 2^w \in \mathbb{N}.$
Then, the value of $( (r \cdot \psi) \mymod p_2 )$ can be calculated as
$$\varphi = \redc(\widetilde{r}, \psi),$$
where the $\redc$ is done for the modulo $p_2.$

@@

The idea for this optimization is due to~\cite{shenanigans}.

Remember the procedure $\mathrm{MontgomeryReduce}(a)$ that we defined in
section~\ref{section:montgomery_intro}.
It requires $$0 \le a < \mu p,$$ where $p$ is our prime modulo,
$\mu = 2^w,$ $w$ is the length of machine word in bits.
In order to compute $\redc(x,y),$ we call $\mathrm{MontgomeryReduce}(xy).$
Normally, we have $0 \le x,y < p.$
By~\eqref{eq:pi_bound}, we also have $p < \mu/2.$
It means that $$a = xy < p \mu / 2.$$

This means that we can call $\mathrm{MontgomeryReduce}(xy)$ if either $x$ or $y$ (but not both)
is in ``unadjusted'' form. Being ``unadjusted'' means being in $[0; 2p-1],$
thus possibly having different representation than the ``canonical'' one for this value
(in $[0; p-1]$).

We can obtain such an ``unadjusted'' value by omitting an adjustment:
\begin{itemize}
    \item After an addition: the unadjusted addition of $x,y \in [0; p-1]$ is just $x + y.$
    \item After a subtraction: the unadjusted subtraction of $x,y \in [0; p-1]$ is $x - y + p.$
    \item After $\redc$: observe that $\mathrm{MontgomeryReduce}$ itself
          performs adjustment as the last step.
\end{itemize}

Note that we use this optimization much more sparingly than we theoretically could, because
allowing non-local propagation of ``unadjustment'' would be a nightmare from the standpoint of
reliability and debuggability: we would need to keep track of what can, and what can not, be
``unadjusted'' at each step.

\section{Prospects}

We see the following possible ways to further improve the performance of decimal multiplication.

\begin{enumerate}
  \item \textbf{Improve cache utilization}.
    \begin{enumerate}
      \item Use breadth-first ordering, perhaps using explicit recursion.
            This can be beneficial even considering the function call overhead,
            as it leads to better cache utilization~\cite{fftw}.
      \item Explore the ``Belgian approach''; see~\cite{belgian}.
      \item Explore using higher-radix transforms (raidx-4, radix-8, etc).
            Note that radix-4 is beneficial for complex FFT because,
            for a standard representation by a pair of floating-point values,
            multiplication by $i$ or $-i$ can be done significantly faster than normal
            multiplication; but finite fields of integers modulo $p$ with standard representation
            lack this property.
            Nevertheless, higher-radix transforms might be of service because they have better cache
            locality~\cite{fftw}.
    \end{enumerate}
  \item \textbf{Additional shenanigans}.
    \begin{enumerate}
      \item It is suggested in \cite{shenanigans, mont_sp} to use primes $p$ such that $p < \mu/4.$
          Then, the Montgomery multiplication can be implemented in such a way that if the inputs
          are in $[0; 2p-1],$ then the outputs are also in $[0; 2p-1],$ and no final conditional
          subtraction is performed.
      \item It is stated in~\cite{shenanigans_presentation} that we could benefit from
          higher-radix transforms by further eliminating reductions from the butterflies.
      \item Explore signed Montgomery reduction, as defined in~\cite{signed_montgomery}.
    \end{enumerate}
  \item \textbf{Smooth the stairs}, possibly utilizing ideas from~\cite{smooth_the_stairs}.
        Currently, we can only perform transforms of lengths $2^k$ and $3 \cdot 2^k.$
        It would be nice to be able to perform transforms of lengths that are in-between.
\end{enumerate}

\section{Benchmark}

\begin{figure}[H]
    \centering
    \begin{gnuplot}[terminal=epslatex, terminaloptions=color]
        set style line 1 lt 1 lw 2 lc rgb '#4682b4' pt -1
        set style line 2 lt 1 lw 2 lc rgb '#ee0000' pt -1

        set xtics axis
        set ytics axis

        set grid xtics lc rgb '#555555' lw 1 lt 0
        set grid ytics lc rgb '#555555' lw 1 lt 0

        set ytics 0.5
        set xtics 1

        set xlabel "$log_2(n)$"
        set ylabel "ratio"

        set autoscale xfix

        plot \
            'benchdata.txt' using 1:2 with lines ls 1 ti "CMOVs on", \
            'benchdata.txt' using 1:3 with lines ls 2 ti "CMOVs off"
    \end{gnuplot}
\end{figure}
